Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-09 csf-2017/paper.tex
added a bit more to LR introduction
Some work on the capability machine section.

\usepackage{cite}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{thmtools}
\usepackage{stmaryrd}
\usepackage{natbib}
\usepackage{url}
\usepackage{array}
\usepackage{arydshln}
\usepackage{ifthen}
\usepackage{ifpdf}
\usepackage{verbatim}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{natbib}
\lstset{
  basicstyle=\ttfamily,
  columns=fullflexible,
  keepspaces=true,
  mathescape
}

@@

\section{A capability machine with local capabilities}
\begin{figure}
  \begin{align*}
    \addr   &\in& \Addrs &\isdef \nats\\
    w &\in&\Words &\isdef \ints + \Caps \\
    \perm   &\in& \Perms &::= \noperm \mid \readonly\mid \readwrite\mid \readwritel\mid \exec\mid \entry\mid \rwx\mid \rwlx\\
    \gl&\in&\Globals & ::= \glob \mid \local \\
     && \Caps  &::= \left\{
                       \begin{multlined}
                         ((\perm,\gl),\start,\addrend,\addr)\mid\\
                         \start\in\Addrs,\addrend \in
                         \Addrs\cup \{\infty\}
                       \end{multlined} \right\}\\
    r       &\in& \RegName&\bnfdef \pcreg\mid r_0\mid r_1\mid\ldots\\
    \reg &\in& \Regs  &\isdef \RegName \rightarrow \Words\\
    m&\in& \Heaps &\isdef \Addrs \rightarrow \Words \\
    \Phi    &\in& \ExecConfs  &\isdef \Regs \times \Heaps \\
    &&\Confs &::= \ExecConfs + \{\failed \} + \{\halted\} \times \Heaps \\
    \ms     &\in& \MemSegments &::= \Addrs \parfun \Words
  \end{align*}
\[
  \begin{array}{rcl}
    \lv    &::=& \refreg{r} \\
    \hv    &::=& \refheap{r}\\
    \rv    &::=& n \mid \lv \\
    i      &::=& 
                 \jmp{\lv} \mid 
                 \jnz{\lv}{\rv} \mid
                 \move{\lv}{\rv} \mid 
                 \load{\lv}{\hv} \mid 
                 \store{\hv}{\rv} \mid  \\
           &   & \plus{\lv}{\rv}{\rv} \mid 
                 \lea{\lv}{\rv} \mid 
                 \restricttwo{\lv}{\rv} \mid 
                 \subseg{\lv}{\rv}{\rv} \mid  \\
           &   & \isptr{\lv}{\rv} \mid 
                 \getp{\lv}{\lv} \mid 
                 \getl{\lv}{\lv} \mid 
                 \getb{\lv}{\lv} \mid
                 \gete{\lv}{\lv} \mid\\
           &   & \geta{\lv}{\lv} \mid 
                 \fail \mid
                 \halt 
  \end{array}
\]
  \caption{The syntax of our capability machine assembly language.}
  \label{fig:syntax}
\end{figure}

@@

We define the syntax of our capability machine in \figurename~\ref{fig:syntax}. We
assume an infinite set of addresses $\Addrs$ and define machine words as either
integers or capabilities of the form $((\perm,\gl),\start,\addrend,\addr)$. Such
a capability represents the authority to execute permissions $\perm$ on the
memory array $[\start,\addrend]$, together with a current address $\addr$ and a
tag $\gl$ which indicates whether the capability is global or local. The
available permissions and their ordering are depicted in
\figurename~\ref{fig:perm-hier}: the permissions include null permission ($\noperm$),
readonly ($\readonly$), read/write ($\readwrite$), read/execute ($\exec$) and
read/write/execute ($\rwx$) permissions. Additionally, there are somewhat
special read/write-local ($\readwritel$), read/write-local/execute ($\rwlx$) and
enter ($\entry$) permissions which we will explain below.

We assume a set of register names $\RegName$. We define register files $\reg$ as
functions mapping register names to words and memories as functions that map
addresses to words. The state of the entire machine is represented as a
configuration that is either a running $\Phi \in \ExecConfs$ containing a memory
and a register file, or a failed or halted state, where the latter keeps track
of the final state of memory.

@@

\subsection{Worlds}
\todo[inline]{Find a good place to describe monotonicity of
  interpretation function.}
\begin{figure}[htbp]
  \centering
  \includegraphics{w11}
  \caption{Relation between local and global capabilities and temporary and permanent regions.}
  \label{fig:cap-world}
\end{figure}
A world is a finite map from region names, modelled as natural numbers, to regions.
We have three types of regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, which describes the protocol governing
the evolution of the memory described by the region, similarly to what
has been used in logical relations for high-level
languages~\cite{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. 
The protocols imposed on memory by permanent regions stay in place
indefinitely. This means that any capability, local or global, can
depend on these protocols. The protocols imposed on memory by temporary
regions can be revoked at certain points during execution,
corresponding to the points when local capabilities can be
revoked. This means that local capabilities safely can depend on the
protocols imposed on memory by temporary regions, whereas global
capabilities cannot, since a global capability may stay around
throughout the execution of a program, but a temporary region may be
revoked.
\todo[inline]{Insert reference to figure.}

@@

The conditions $\entryCond{}$ and $\execCond{}$ are very similar and
differ only slightly. Executable capabilities can be updated to point
anywhere in memory, so they must be capability safe as a program
counter no matter where they point in memory. In contrast, enter
\lars{``capability safe as a program counter'' is a bit hard to
  understand - why not say that this is a capability one can jump to ?}
\lau{ When jumping to a capability, it becomes the program counter. A program counter is capability safe when it is the $\stder$ relation. When I say ``capability safe as a program counter'' I take it to mean being in the $\stder$-relation. }
capabilities are opaque and must be used from the address they point
to. They change permission when jumped to, so we require them to be
capability safe as a program counter when the permission is changed to
read/execute.
In both conditions, the capability must be capability safe as a
program counter in certain future worlds. If the capability is global,
then it should be capability safe as a program counter in any private
future world. This means that it can only depend on permanent regions
because temporary regions may be revoked in a private future
world. On the other hand, local capabilities must only be capability
safe in public future worlds because they may depend on temporary
regions, so we cannot expect them to function correctly as program
counters if temporary regions are revoked.

>>>>

\usepackage{cite}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{thmtools}
\usepackage{stmaryrd}
\usepackage{natbib}
\usepackage{url}
\usepackage{array}
\usepackage{arydshln}
\usepackage{ifthen}
\usepackage{ifpdf}
\usepackage{verbatim}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{multicol}
\usepackage{natbib}
\lstset{
  basicstyle=\ttfamily,
  columns=fullflexible,
  keepspaces=true,
  mathescape
}

@@

\section{A capability machine with local capabilities}
In this paper, we work with a formal capability machine that has all the
characteristics of real capability machines, as well as local capabilities much
like CHERI's, but is otherwise kept as simple as possible. It is inspired by
both the M-Machine~\cite{Carter:1994:HSF:195473.195579} and
CHERI~\cite{Watson2015Cheri}. To avoid getting bogged down in uninteresting
detail, we assume an infinite address space and unbounded integers.

\begin{figure}
  \begin{align*}
    \addr   &\in& \Addrs &\isdef \nats\\
    w &\in&\Words &\isdef \ints + \Caps \\
    \perm   &\in& \Perms &::= \noperm \mid \readonly\mid \readwrite\mid \readwritel\mid \exec\mid \entry\mid \rwx\mid \rwlx\\
    \gl&\in&\Globals & ::= \glob \mid \local \\
     && \Caps  &::= \left\{
                       \begin{multlined}
                         ((\perm,\gl),\start,\addrend,\addr)\mid\\
                         \start\in\Addrs,\addrend \in
                         \Addrs\cup \{\infty\}
                       \end{multlined} \right\}\\
    r       &\in& \RegName&\bnfdef \pcreg\mid r_0\mid r_1\mid\ldots\\
    \reg &\in& \Regs  &\isdef \RegName \rightarrow \Words\\
    m&\in& \Heaps &\isdef \Addrs \rightarrow \Words \\
    \Phi    &\in& \ExecConfs  &\isdef \Regs \times \Heaps \\
    &&\Confs &::= \ExecConfs + \{\failed \} + \{\halted\} \times \Heaps \\
    \ms     &\in& \MemSegments &::= \Addrs \parfun \Words
  \end{align*}
  \begin{equation*}
  \begin{array}{rcl}
    \lv    &::=& \refreg{r} \\
    \hv    &::=& \refheap{r}\\
    \rv    &::=& n \mid \lv \\
    i      &::=& 
                 \jmp{\lv} \mid 
                 \jnz{\lv}{\rv} \mid
                 \move{\lv}{\rv} \mid 
                 \load{\lv}{\hv} \mid 
                 \store{\hv}{\rv} \mid  \\
           &   & \plus{\lv}{\rv}{\rv} \mid 
                 \lea{\lv}{\rv} \mid 
                 \restricttwo{\lv}{\rv} \mid 
                 \subseg{\lv}{\rv}{\rv} \mid  \\
           &   & \isptr{\lv}{\rv} \mid 
                 \getp{\lv}{\lv} \mid 
                 \getl{\lv}{\lv} \mid 
                 \getb{\lv}{\lv} \mid
                 \gete{\lv}{\lv} \mid\\
           &   & \geta{\lv}{\lv} \mid 
                 \fail \mid
                 \halt 
  \end{array}
\end{equation*}
  \caption{The syntax of our capability machine assembly language.}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure*}
  \centering
  \begin{equation*}
    \Phi  \rightarrow
    \begin{cases}
      \sem{\decode(\memheap(\addr))}(\Phi) & \text{if $\memreg(\pcreg) = \stdcap$ and $\start \leq \addr \leq \addrend$ and $\perm \in \{ \exec,\rwx, \rwlx \}$ } \\
      \failed                                 & \text{otherwise}
    \end{cases}
  \end{equation*}%
  \begin{align*}
    \updatePcPerm{w} &=
                       \begin{cases}
                         ((\exec,\gl),\start,\addrend,\addr) & \text{if $w = ((\entry,\gl),\start,\addrend,\addr)$}\\
                         w & \text{otherwise}
                       \end{cases}\\
    \stdUpdatePc{\Phi} &=
                         \begin{cases}
                           \updateReg{\pcreg}{\var{newPc}} & \arraycolsep=0pt
                           \begin{array}[t]{l}
                             \text{if $\memreg(\pcreg) = \stdcap$}\\
                             \quad\text{and $\var{newPc} = ((\perm,\gl),\start,\addrend,\addr + 1)$}\\
                           \end{array}
                         \end{cases}
  \end{align*}

  \begin{tabular}{|c|p{3.7cm}|p{10.5cm}|}
    \hline
    $i$&$\sem{i}(\Phi)$&Conditions\\
    \hline
    $\fail$&$\failed$&\\
    \hline
    $\halt$&$(\halted,\memheap)$&\\
    \hline
    $\move{\refreg{r_1}}{\rv}$& $\begin{multlined}
      \stdUpdatePc{}(\updateReg{r_1}{\\
          \memreg(\rv)}) \end{multlined}$&\\
    \hline
    $\jmp{\lv}$&$\begin{multlined}
                   \updateReg{\pcreg}{\\\updatePcPerm{\memreg(\lv)}}
                 \end{multlined}$&\\
    \hline
    $\load{\refreg{r_1}}{\refheap{r_2}}$&$\updateReg{r_1}{\memheap(\addr)}$&$\memreg(r_2) = \stdcap$ and $\perm \in \{ \rwx, \rwlx, \exec, \readwrite, \readwritel, \readonly \}$  and  $\start \leq \addr \leq \addrend$\\
    \hline
    $\store{\refheap{r_1}}{\refreg{r_2}}$&$\stdUpdatePc{\updateHeap{\addr}{\var{w}}}$&$\memreg(r_1) = \stdcap$ and $\perm \in \{ \rwx, \rwlx, \readwrite, \readwritel\}$  and  $\start \leq \addr \leq \addrend$ and $\var{w} = \memreg(r_2)$
                                                                                       and if  $\var{w} = ((\_,\local),\_,\_,\_)$, then  $\perm \in \{\rwlx,\readwritel \}$\\
    \hline
    $\restricttwo{\refreg{r}}{\rv}$&$\begin{multlined}
                                       \stdUpdatePc{}(\updateReg{r}{\\(\permp',\start,\addrend,\addr)})
                                     \end{multlined}$  &
                                                      $\memreg(r) = \stdcap[\permp]$ and $\permp' = \decodePermPair{\memreg(\rv)}$
                                                      and $\permp'\sqsubseteq \permp$\\
    \hline
    $\geta{\refreg{r_1}}{\refreg{r_2}}$ & $\stdUpdatePc{\updateReg{r_1}{\addr}}$ &
                                                $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$\\
    \hline
    \multicolumn{3}{|c|}{$\cdots$}\\
    \hline
    \_&$\failed$&otherwise\\
    \hline
  \end{tabular}
  \caption{An excerpt from the operational semantics}
  \label{fig:op-sem}
\end{figure*}

We define the syntax of our capability machine in Figure~\ref{fig:syntax}. We
assume an infinite set of addresses $\Addrs$ and define machine words as either
integers or capabilities of the form $((\perm,\gl),\start,\addrend,\addr)$. Such
a capability represents the authority to execute permissions $\perm$ on the
memory array $[\start,\addrend]$, together with a current address $\addr$ and a
tag $\gl$ which indicates whether the capability is global or local. The
available permissions and their ordering are depicted in
Figure~\ref{fig:perm-hier}: the permissions include null permission ($\noperm$),
readonly ($\readonly$), read/write ($\readwrite$), read/execute ($\exec$) and
read/write/execute ($\rwx$) permissions. Additionally, there are three special
read/write-local ($\readwritel$), read/write-local/execute ($\rwlx$) and enter
($\entry$) permissions which we will explain below.

We assume a set of register names $\RegName$. We define register files $\reg$ as
functions mapping register names to words and memories as functions that map
addresses to words. The state of the entire machine is represented as a
configuration that is either a running $\Phi \in \ExecConfs$ containing a memory
and a register file, or a failed or halted state, where the latter keeps hold
of the final state of memory. We also define memory segments $\ms$:
parts of memory represented as partial functions mapping addresses to words.

The instruction set of our machine is rather basic.  Instructions $i$ include
jumps ($\jmp{}$), conditional jumps ($\jnz{}{}$), a move instruction that copies
a word from one register to another ($\move{}{}$), load and store instructions
for reading from and writing to memory ($\load{}{}$ and $\store{}{}$)

@@

\subsection{Worlds}
\todo[inline]{Find a good place to describe monotonicity of
  interpretation function.}
\begin{figure}[htbp]
  \centering
  \includegraphics{w11}
  \caption{Relation between local and global capabilities and temporary and permanent regions.}
  \label{fig:cap-world}
\end{figure}
A world is a finite map from region names, modelled as natural numbers, to regions.
We have three types of regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, which describes the protocol governing
the evolution of the memory described by the region, similarly to what
has been used in logical relations for high-level
languages~\cite{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. 
The protocols imposed on memory by permanent regions stay in place
indefinitely. This means that any capability, local or global, can
depend on these protocols. The protocols imposed on memory by temporary
regions can be revoked at certain points during execution,
corresponding to the points when local capabilities can be
revoked. This means that local capabilities safely can depend on the
protocols imposed on memory by temporary regions, whereas global
capabilities cannot, since a global capability may stay around
throughout the execution of a program, but a temporary region may be
revoked.

@@

The conditions $\entryCond{}$ and $\execCond{}$ are very similar and
differ only slightly. Executable capabilities can be updated to point
anywhere in memory, so they must be capability safe as a program
counter no matter where they point in memory. In contrast, enter
\lars{``capability safe as a program counter'' is a bit hard to
  understand - why not say that this is a capability one can jump to ?}
\lau{ When jumping to a capability, it becomes the program counter. A program counter is capability safe when it is the $\stder$ relation. When I say ``capability safe as a program counter'' I take it to mean being in the $\stder$-relation. }
capabilities are opaque and must be used from the address they point
to. They change permission when jumped to, so we require them to be
capability safe as a program counter when the permission is changed to
read-execute.
In both conditions, the capability must be capability safe as a
program counter in certain future worlds. If the capability is global,
then it should be capability safe as a program counter in any private
future world. This means that it can only depend on permanent regions
because temporary regions may be revoked in a private future
world. On the other hand, local capabilities must only be capability
safe in public future worlds because they may depend on temporary
regions, so we cannot expect them to function correctly as program
counters if temporary regions are revoked.
